Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2025-01-14 paper/main.tex
slight re-framing of Fig 6 across-lecture results

Finally, when we estimated participants' knowledge for each question about one
lecture using their performance on questions (from the same quiz) about the
\textit{other} lecture, we observed a somewhat different pattern of results.
Here we found that before viewing either lecture (i.e., on Quiz~1),
participants' abilities to answer \textit{Four Fundamental Forces}-related
questions could not be predicted from their responses to \textit{Birth of
Stars}-related questions ($OR = 1.896,\ 95\%\ \textnormal{CI} = [0.419,\
9.088],\ \lambda_{LR} = 0.712,\ p = 0.404$), nor could their abilities to
answer \textit{Birth of Stars}-related questions be predicted from their
responses to \textit{Four Fundamental Forces}-related questions ($OR = 1.522,\
95\%\ \textnormal{CI} = [0.332,\ 6.835],\ \lambda_{LR} = 0.286,\ p = 0.611$).
Similarly, we found that participants' performance on questions about either
lecture could not be predicted given their responses to questions about the
other lecture after viewing \textit{Four Fundamental Forces} but before viewing
\textit{Birth of Stars} (i.e., on Quiz~2; \textit{Four Fundamental Forces}
questions given \textit{Birth of Stars} questions: $OR = 3.49,\ 95\%\
\textnormal{CI} = [0.739,\ 12.849],\ \lambda_{LR} = 3.266,\ p = 0.083$;
\textit{Birth of Stars} questions given \textit{Four Fundamental Forces}
questions: $OR = 2.199,\ 95\%\ \textnormal{CI} = [0.711,\ 5.623],\
\lambda_{LR} = 2.304,\ p = 0.141$). Only after viewing \textit{both} lectures
(i.e., on Quiz~3) did these across-lecture knowledge estimates reliably predict
participants' success on individual quiz questions (\textit{Four Fundamental
Forces} questions given \textit{Birth of Stars} questions: $OR = 11.294,\
95\%\ \textnormal{CI} = [1.375,\ 47.744],\ \lambda_{LR} = 10.396,\ p < 0.001$;
\textit{Birth of Stars} questions given \textit{Four Fundamental Forces}
questions: $OR = 7.302,\ 95\%\ \textnormal{CI} = [1.077,\ 44.879],\
\lambda_{LR} = 4.708,\ p = 0.038$). Taken together, these results suggest that
our knowledge estimates can be used to predict participants' success across
content areas once they have received some training on both the content about
which their knowledge is estimated and the content used to construct these
estimates.

>>>>

Finally, we estimated participants' knowledge for each question about one
lecture using their performance on questions (from the same quiz) about the
\textit{other} lecture. This is an especially stringent test of our approach.
Our primary assumption in building our knowledge estimates is that knowledge
about a given concept is similar to knowledge about other concepts that are
nearby in the embedding space. However, our analyses in Figure~\ref{fig:topics}
and Supplementary Figure~\topicWeights~show that the embeddings of content from
the two lectures are largely distinct. Therefore any predictive power of the
knowledge estimates must overcome large distances in the embedding space. To
put this in concrete terms, this test requires predicting participants'
performance on individual highly specific questions about the formation of
stars, using each participants' responses to just five multiple choice
questions about the fundamental forces of the universe (and vice versa).

We found that, before viewing either lecture (i.e., on Quiz~1), participants'
abilities to answer \textit{Four Fundamental Forces}-related questions could
not be predicted from their responses to \textit{Birth of Stars}-related
questions ($OR = 1.896,\ 95\%\ \textnormal{CI} = [0.419,\ 9.088],\ \lambda_{LR}
= 0.712,\ p = 0.404$), nor could their abilities to answer \textit{Birth of
Stars}-related questions be predicted from their responses to \textit{Four
Fundamental Forces}-related questions ($OR = 1.522,\ 95\%\ \textnormal{CI} =
[0.332,\ 6.835],\ \lambda_{LR} = 0.286,\ p = 0.611$). Similarly, we found that
participants' performance on questions about either lecture could not be
predicted given their responses to questions about the other lecture after
viewing \textit{Four Fundamental Forces} but before viewing \textit{Birth of
Stars} (i.e., on Quiz~2; \textit{Four Fundamental Forces} questions given
\textit{Birth of Stars} questions: $OR = 3.49,\ 95\%\ \textnormal{CI} =
[0.739,\ 12.849],\ \lambda_{LR} = 3.266,\ p = 0.083$; \textit{Birth of Stars}
questions given \textit{Four Fundamental Forces} questions: $OR = 2.199,\ 95\%\
\textnormal{CI} = [0.711,\ 5.623],\ \lambda_{LR} = 2.304,\ p = 0.141$). Only
after viewing \textit{both} lectures (i.e., on Quiz~3) did these across-lecture
knowledge estimates reliably predict participants' success on individual quiz
questions (\textit{Four Fundamental Forces} questions given \textit{Birth of
Stars} questions: $OR = 11.294,\ 95\%\ \textnormal{CI} = [1.375,\ 47.744],\
\lambda_{LR} = 10.396,\ p < 0.001$; \textit{Birth of Stars} questions given
\textit{Four Fundamental Forces} questions: $OR = 7.302,\ 95\%\ \textnormal{CI}
= [1.077,\ 44.879],\ \lambda_{LR} = 4.708,\ p = 0.038$). Taken together, our
results suggest that our ability to form estimates solely across different
content areas is more limited than our ability to form estimates that
incorporate responses to questions across both content areas (as in
Fig.~\ref{fig:predictions}, ``All questions'') or within a single content area (as
in Fig.~\ref{fig:predictions}, ``Within-lecture'').  However, if participants have recently
received some training on both content areas, the knowledge estimates appear to be informative
even across content areas.

We speculate that these ``Across-lecture'' results might relate to some of our
earlier work on the nature of semantic representations~\citep{MannKaha12}. In
that work, we asked whether semantic similarities could be captured through
behavioral measures, even if participants' ``true'' internal representations
differed from the embeddings used to \textit{characterize} participants'
behaviors. We found that mismatches between someone's internal representation
of a set of concepts and the representation used to characterize their
behaviors can lead to underestimates of how semantically driven their behaviors
are. Along similar lines, we suspect that in our current study, participants'
conceptual representations may initially differ from the representations
learned by our topic model. (Although the topic models are still
\textit{related} to participants' initial internal representations; otherwise
we would have found that knowledge estimates derived from Quiz 1 and 2
responses would have no predictive power in the other tests we conducted.)
After watching both lectures, however, participants' internal representations
may become more aligned with the embeddings used to estimate their knowledge
(since those embeddings were trained on the lecture transcripts). This could
help explain why the knowledge estimates derived from Quizzes 1 and 2 (before
both lectures had been watched) do not reliably predict performance across
content areas, whereas estiamtes derived from Quiz 3 \textit{do} reliably
predict performance across content areas.
